Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-10 demystify.tex
updates

\paragraph{Self-written letters are bad} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \textbf{very little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

\paragraph{Generic letters are bad} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak. This does not mean the professor is not good or does not care about you, but they just do not know you well enough to write a strong letter.
So it might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

\paragraph{Waiving Your Right}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{Publications} Concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has been successfully involved in research.  Applicants admitted to top schools often have multiple first-authored papers at top places.

>>>>

\subsection{Generic Letters are Bad} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.

This does not mean the professor is not good or does not care about you, but they just do not know you well enough to write a strong letter.
So it might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

\subsection{Self-written Letters are Bad} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \textbf{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

Most well-known profs. would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). However, many international students do not know this (e.g., normal at their university) and might be in a situation they do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer or relatively unknown in the field.  In either case, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

\subsubsection{Letters from Admin people}  
The ``department head'', ``dean'' or ``director'' at your university might not be well-known to adcom members (sorry, we likely recognized researchers, not admin people), so it's fine to to have "less well-known" people write for you as long as they can write a good letter about your research ability).

\subsection{Waiving Your Right}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{Publications} The most concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the you has been successfully involved in research. Publications are never required for application, however given the competitiveness of the process, they can significantly strengthen your application and are becoming the norm, e.g., those admitted to top schools often have multiple first-authored papers at top places.
